---
title: "Predictive Modelling for Wages"
format: html
editor: visual
---

## 1. Setup and Libraries

We use `tidyverse` for data manipulation, `caret` for preprocessing, `xgboost` as our primary model, and `DALEX` for explainability.

```{r setup, message=FALSE, warning=FALSE}
packages <- c("tidyverse", "caret", "xgboost",
              "DALEX", "ggplot2", "corrplot",
              "Metrics", "gridExtra")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

library(tidyverse)
library(caret)
library(xgboost)
library(DALEX)
library(ggplot2)
library(corrplot)
library(Metrics)
library(gridExtra)

load("data_wage.Rdata")
```

------------------------------------------------------------------------

## 2. Exploratory Data Analysis

```{r eda, message=FALSE, warning=FALSE}
dim(data)
```

Chech for missing Data.

```{r eda-missing}
missing_summary <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(missing_pct = round(missing_count / nrow(data) * 100, 1)) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_pct))
print(missing_summary)
```

We create a histogram of raw wage values. The second plot are for wages where they are really high and with the help of Log transformation we can now have a better understanding.

```{r eda-wage-dist, fig.width=10, fig.height=4}
p1 <- ggplot(data, aes(x = wage)) +
  geom_histogram(fill = "steelblue", bins = 40, alpha = 0.8) +
  labs(title = "Distribution of Wage", x = "Wage", y = "Count") +
  theme_minimal()

p2 <- data %>%
  filter(!is.na(wage), wage > 0) %>%
  ggplot(aes(x = log1p(wage))) +
  geom_histogram(fill = "darkorange", bins = 40, alpha = 0.8) +
  scale_x_continuous(
    breaks = log1p(c(1000, 10000, 50000, 100000, 200000, 500000)),
    labels = c("$1k", "$10k", "$50k", "$100k", "$200k", "$500k")
  ) +
  labs(title = "Distribution of log(Wage + 1)", 
       x = "Wage (USD)", y = "Count") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The distribution is strongly right-skewed with notable outliers. We apply a log transformation to reduce skew and improve model performance.

```{r eda-categorical, fig.width=10, fig.height=5}
if ("gender" %in% colnames(data)) {
  p3 <- data %>%
  filter(!is.na(gender), !is.na(wage), wage > 0) %>%
  ggplot(aes(x = gender, y = wage, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Wage by Gender (log scale)", x = "Gender", y = "Wage (log scale)") +
  theme_minimal() + theme(legend.position = "none")
  print(p3)
}

if ("education" %in% colnames(data)) {
 p4 <- data %>%
  filter(!is.na(education), !is.na(wage), wage > 0) %>%
  ggplot(aes(x = reorder(education, wage, median), y = wage, fill = education)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  coord_flip() +
  labs(title = "Wage by Education Level (log scale)", x = "Education", y = "Wage (log scale)") +
  theme_minimal() + theme(legend.position = "none")
  print(p4)
}

if ("job_role" %in% colnames(data)) {
  p5 <- data %>%
    filter(!is.na(job_role), !is.na(wage)) %>%
    group_by(job_role) %>%
    summarise(median_wage = median(wage, na.rm = TRUE)) %>%
    ggplot(aes(x = reorder(job_role, median_wage), y = median_wage, fill = median_wage)) +
    geom_col(alpha = 0.85) +
    coord_flip() +
    scale_fill_viridis_c() +
    labs(title = "Median Wage by Job Role", x = "Job Role", y = "Median Wage") +
    theme_minimal() + theme(legend.position = "none")
  print(p5)
}
```

The wage distribution shows slight variation across gender groups, however the differences are too small to draw any meaningful conclusions. Regarding education level, a clear positive relationship is visible, higher academic degrees are consistently associated with higher wages. Finally, looking at wage by job role, Chief Officers, Principal Investigators, and Managers stand out as the highest earning roles, earning significantly more than other positions.

```{r eda-numeric, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
# 1. Top programming languages by usage rate
prog_cols <- data %>%
  select(starts_with("Programming_")) %>%
  select(-Programming_language_used_most_often) %>%
  summarise(across(everything(), ~ mean(. == 1, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "language", values_to = "usage_rate") %>%
  mutate(language = str_remove(language, "Programming_")) %>%
  arrange(desc(usage_rate))

p1 <- ggplot(prog_cols, aes(x = reorder(language, usage_rate), y = usage_rate)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Programming Language Usage", x = NULL, y = "% of Respondents") +
  theme_minimal()

# 2. ML framework usage rate
ml_cols <- data %>%
  select(starts_with("ML_framework_")) %>%
  summarise(across(everything(), ~ mean(. == 1, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "framework", values_to = "usage_rate") %>%
  mutate(framework = str_remove(framework, "ML_framework_")) %>%
  arrange(desc(usage_rate))

p2 <- ggplot(ml_cols, aes(x = reorder(framework, usage_rate), y = usage_rate)) +
  geom_col(fill = "darkorange", alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "ML Framework Usage", x = NULL, y = "% of Respondents") +
  theme_minimal()

# 3. Wage by programming language (Python vs R vs SQL users)
lang_wage <- data %>%
  filter(!is.na(wage)) %>%
  mutate(
    uses_python = ifelse(Programming_Python == 1, "Python", "No Python"),
    uses_r      = ifelse(Programming_R == 1,      "R",      "No R"),
    uses_sql    = ifelse(Programming_SQL == 1,     "SQL",    "No SQL")
  ) %>%
  pivot_longer(cols = c(uses_python, uses_r, uses_sql),
               names_to = "lang_type", values_to = "uses_lang") %>%
  filter(!is.na(uses_lang))

p3 <- ggplot(lang_wage, aes(x = uses_lang, y = wage, fill = uses_lang)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ lang_type, scales = "free_x") +
  labs(title = "Wage by Key Programming Language Usage (log scale)",
       x = NULL, y = "Wage (log scale)") +
  theme_minimal() +
  theme(legend.position = "none")

# 4. Wage by cloud platform usage
cloud_wage <- data %>%
  filter(!is.na(wage)) %>%
  mutate(
    uses_aws   = ifelse(cloud_Amazon.Web.Services..AWS.       == 1, "AWS",   "No AWS"),
    uses_gcp   = ifelse(cloud_Google.Cloud.Platform..GCP.     == 1, "GCP",   "No GCP"),
    uses_azure = ifelse(cloud_Microsoft.Azure                  == 1, "Azure", "No Azure")
  ) %>%
  pivot_longer(cols = c(uses_aws, uses_gcp, uses_azure),
               names_to = "cloud_type", values_to = "uses_cloud") %>%
  filter(!is.na(uses_cloud))

p4 <- ggplot(cloud_wage, aes(x = uses_cloud, y = wage, fill = uses_cloud)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ cloud_type, scales = "free_x") +
  labs(title = "Wage by Cloud Platform Usage (log scale)",
       x = NULL, y = "Wage (log scale)") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The relationship between programming language usage and wages is not immediately clear, making it difficult to draw strong conclusions from the data. However, cloud platform usage tells a different story, there is a notable difference in wages between users and non-users across all three platforms. In particular, AWS users appear to earn significantly more than non-users, suggesting that cloud platform experience, especially AWS, is associated with higher compensation. This may reflect the fact that cloud skills are in high demand in the industry, commanding a wage premium.

------------------------------------------------------------------------

## 3. Data Transformation and Cleaning

### 3.1 Feature Selection Rationale

We select features based on three criteria:

1.  **Domain relevance**: Variables that intuitively drive wages (experience, education, skills, role, country)
2.  **Data quality**: Dropping columns with \>40% missing values
3.  **Avoiding data leakage**: Excluding any variable that directly encodes wage

Wages are not internationally uniform, compensation in prosperous economies is significantly higher than in developing countries, making country of employment one of the most important variables in our transformation and modelling process. As part of our data quality checks, we also screened for missing values. We applied a 40% threshold, meaning any variable with more than 40% missing values was removed. This ensures we only retain features that carry meaningful information. Finally, to prevent data leakage, we carefully excluded any variables that directly encode wage. Including such variables would artificially inflate model performance, as the model would essentially be using the answer to predict the answer, producing misleadingly optimistic results that would not generalise to new data.

```{r cleaning, message=FALSE, warning=FALSE}
# Step 1: Remove rows where wage is missing
data_clean <- data %>% filter(!is.na(wage))
cat("Rows after removing missing wage:", nrow(data_clean), "\n")

# Step 2: Drop columns with >40% missing values
missing_threshold <- 0.40
keep_cols <- colnames(data_clean)[colMeans(is.na(data_clean)) <= missing_threshold]
data_clean <- data_clean %>% select(all_of(keep_cols))
cat("Columns remaining after missing filter:", ncol(data_clean), "\n")

# Step 3: Log-transform the target (wage is right-skewed)
data_clean <- data_clean %>%
  mutate(log_wage = log1p(wage)) %>%
  select(-wage)

# Step 4: Handle remaining NAs
data_clean <- data_clean %>%
  mutate(across(where(is.character), ~ replace_na(., "Unknown"))) %>%
  mutate(across(where(is.factor),   ~ fct_explicit_na(., na_level = "Unknown"))) %>%
  mutate(across(where(is.numeric),  ~ replace_na(., median(., na.rm = TRUE))))

cat("Final clean dataset dimensions:", dim(data_clean), "\n")
```

```{r encoding}
# Step 5: Encode character columns as factors
cat_vars <- data_clean %>% select(where(is.character)) %>% colnames()
data_model <- data_clean %>%
  mutate(across(all_of(cat_vars), as.factor))
```

### 3.2 Train / Test Split

We apply an 80/20 train/test split, allocating 80% of the data for training and 20% for evaluation. Stratification on log_wage ensures that both sets have a representative wage distribution, preventing the model from being evaluated on an unrepresentative sample. This split ratio was selected after experimenting with alternative configurations, including 70/30 and 90/10 splits, and found to offer the best balance between having sufficient training data for the model to learn from and enough test data to produce reliable performance estimates.

```{r split}
set.seed(42)

train_index <- createDataPartition(data_model$log_wage, p = 0.8, list = FALSE)
train_data  <- data_model[train_index, ]
test_data   <- data_model[-train_index, ]

cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:    ", nrow(test_data),  "\n")
```

### 3.3 One-Hot Encoding for XGBoost

XGBoost requires fully numeric input. We one-hot encode all factor columns using `dummyVars`, resulting in 260+ binary features.

```{r ohe}
dummies   <- dummyVars(log_wage ~ ., data = train_data)
train_xgb <- predict(dummies, newdata = train_data) %>% as.data.frame()
train_xgb$log_wage <- train_data$log_wage

test_xgb  <- predict(dummies, newdata = test_data) %>% as.data.frame()
test_xgb$log_wage  <- test_data$log_wage

# Clean column names
names(train_xgb) <- make.names(names(train_xgb), unique = TRUE)
names(test_xgb)  <- make.names(names(test_xgb),  unique = TRUE)

# XGBoost matrices
X_train <- as.matrix(train_xgb %>% select(-log_wage))
y_train <- train_xgb$log_wage
X_test  <- as.matrix(test_xgb  %>% select(-log_wage))
y_test  <- test_xgb$log_wage

dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

cat("Total features after encoding:", ncol(X_train), "\n")
```

------------------------------------------------------------------------

## 4. Model: XGBoost

### 4.1 Why XGBoost?

We chose XGBoost (Extreme Gradient Boosting) as our model for the following reasons:

-   **Best-in-class for tabular data**: XGBoost consistently outperforms other algorithms on structured datasets with mixed feature types
-   **Handles sparsity natively**: Our dataset has 260+ binary columns (programming languages, tools, activities) — XGBoost is specifically optimised for sparse inputs
-   **Gradient boosting**: Builds trees sequentially, each correcting the errors of the previous one, making it highly accurate
-   **Built-in regularisation**: `lambda`, `alpha`, `gamma`, and `min_child_weight` parameters prevent overfitting on noisy survey data
-   **SHAP compatibility**: Native SHAP support provides state-of-the-art explainability

### 4.2 Hyperparameter Tuning

We use `xgb.cv` with 5-fold cross-validation to find the optimal number of boosting rounds, and tune key parameters to maximise R².

**Parameters explained:**

-   `eta` (0.05): Lower learning rate — each tree contributes less, requiring more rounds but producing a more robust model
-   `max_depth` (5): Tree depth — controls model complexity; 5 balances expressiveness and overfitting
-   `subsample` (0.8): Each tree sees 80% of training rows — adds randomness, reduces overfitting
-   `colsample_bytree` (0.7): Each tree sees 70% of features — forces diversity across trees
-   `min_child_weight` (5): Minimum observations in a leaf — prevents the model from learning noise in small subgroups
-   `gamma` (0.1): Minimum loss reduction to make a split — acts as pruning, improves generalisation

```{r xgboost-tune, message=FALSE, warning=FALSE, cache=TRUE}
# Grid of parameter combinations to try
param_grid <- list(
  list(eta=0.05, max_depth=4, min_child_weight=3,  gamma=0.0, subsample=0.8, colsample_bytree=0.7),
  list(eta=0.05, max_depth=5, min_child_weight=3,  gamma=0.0, subsample=0.8, colsample_bytree=0.7),
  list(eta=0.05, max_depth=6, min_child_weight=3,  gamma=0.0, subsample=0.8, colsample_bytree=0.8),
  list(eta=0.05, max_depth=5, min_child_weight=1,  gamma=0.0, subsample=0.9, colsample_bytree=0.8),
  list(eta=0.05, max_depth=6, min_child_weight=1,  gamma=0.1, subsample=0.8, colsample_bytree=0.6),
  list(eta=0.03, max_depth=5, min_child_weight=3,  gamma=0.0, subsample=0.8, colsample_bytree=0.7),
  list(eta=0.03, max_depth=6, min_child_weight=1,  gamma=0.0, subsample=0.9, colsample_bytree=0.8)
)

best_cv_rmse <- Inf
best_params  <- NULL
best_rounds  <- NULL

set.seed(42)
for (i in seq_along(param_grid)) {
  p <- param_grid[[i]]
  p$objective   <- "reg:squarederror"
  p$eval_metric <- "rmse"
  p$lambda      <- 1.0
  p$alpha       <- 0.0

  cv <- xgb.cv(
    data    = dtrain,
    nrounds = 600,
    nfold   = 5,
    params  = p,
    verbose = FALSE
  )

  min_rmse   <- min(cv$evaluation_log$test_rmse_mean)
  best_round <- which.min(cv$evaluation_log$test_rmse_mean)

  cat(sprintf("Config %d | depth=%d eta=%.2f mcw=%d | CV RMSE: %.4f (round %d)\n",
              i, p$max_depth, p$eta, p$min_child_weight, min_rmse, best_round))

  if (min_rmse < best_cv_rmse) {
    best_cv_rmse <- min_rmse
    best_params  <- p
    best_rounds  <- best_round
  }
}

cat(sprintf("\nBest CV RMSE: %.4f at round %d\n", best_cv_rmse, best_rounds))

# Train final model with best params
set.seed(42)
xgb_model <- xgb.train(
  data    = dtrain,
  nrounds = best_rounds,
  params  = best_params,
  verbose = 0
)
```

### 4.4 Model Evaluation

```{r xgboost-eval}
# Predictions
xgb_preds_log  <- predict(xgb_model, dtest)
xgb_preds_orig <- expm1(xgb_preds_log)
actual_orig    <- expm1(y_test)

# Metrics
xgb_rmse_log  <- rmse(y_test,     xgb_preds_log)
xgb_rmse_orig <- rmse(actual_orig, xgb_preds_orig)
xgb_mae_orig  <- mae(actual_orig,  xgb_preds_orig)
xgb_r2        <- cor(y_test,      xgb_preds_log)^2

cat("=== XGBoost Test Set Performance ===\n")
cat(sprintf("R²:              %.4f\n", xgb_r2))
cat(sprintf("RMSE (log):      %.4f\n", xgb_rmse_log))
cat(sprintf("RMSE (USD):      %.1f\n", xgb_rmse_orig))
cat(sprintf("MAE  (USD):      %.1f\n", xgb_mae_orig))
```

Initially, the XGBoost model produced a lower than expected R² of approximately 0.33. During our investigation we discovered that excluding wages equal to zero nearly doubled the model's accuracy, bringing R² to approximately 0.65. After careful consideration however, we decided to retain the zero-wage observations in our final model. While their removal improves performance metrics, excluding them introduces a selection bias, students, unemployed respondents, and those who chose not to disclose their income are a real and meaningful part of the population. Removing them would make the model less representative of the full workforce and could lead to misleading predictions for profiles that resemble these groups.

```{r eval-plots, fig.width=10, fig.height=4}
residuals_df <- data.frame(
  actual    = actual_orig,
  predicted = xgb_preds_orig,
  residual  = actual_orig - xgb_preds_orig
)

# Actual vs Predicted
p1 <- ggplot(residuals_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue", size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted Wage",
       x = "Actual Wage (USD)", y = "Predicted Wage (USD)") +
  theme_minimal()

# Residuals vs Predicted
p2 <- ggplot(residuals_df, aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.3, color = "darkorange", size = 1) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Predicted",
       x = "Predicted Wage (USD)", y = "Residual") +
  theme_minimal()

# Residual distribution
p3 <- ggplot(residuals_df, aes(x = residual)) +
  geom_histogram(fill = "steelblue", bins = 40, alpha = 0.8) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Distribution of Residuals",
       x = "Residual", y = "Count") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

To evaluate model performance visually we produce three diagnostic plots. The first plot shows Actual vs Predicted Wage, in a perfect model all points would fall exactly on the red dashed line. Points scattered around it indicate prediction errors, with points above the line meaning the model underestimated and points below meaning it overestimated. The second plot shows Residuals vs Predicted, residuals are the difference between actual and predicted wages. Ideally these should be randomly scattered around zero with no clear pattern. A pattern would suggest the model is systematically wrong for certain wage ranges. The third plot shows the Distribution of Residuals, ideally this should be a bell curve centered at zero, meaning errors are symmetric and the model is not consistently over or under predicting. In our case the residuals are approximately centered at zero, which is a good sign, though the spread reflects the inherent difficulty of predicting wages from survey data.

**Performance discussion:**

The model achieves an R² of approximately 0.31–0.40 on this dataset. While this may seem modest, it is expected and acceptable given:

-   **Bucketed variables**: Age and experience are ranges (e.g. 25–29, 1–2 years), not exact values — this loses precision
-   **Self-reported global survey**: Wages vary enormously across countries and roles; a single model must capture all this diversity
-   **Inherent noise**: Survey respondents may approximate or misreport salaries
-   **No continuous numeric predictors**: Almost all features are categorical, limiting the model's ability to capture fine-grained patterns

An R² of 0.31+ on a noisy global wage survey is consistent with published research on similar datasets.

------------------------------------------------------------------------

## 5. Model Explainability

### 5.1 Feature Importance

XGBoost's built-in importance measures show which features contributed most to splits across all trees. We show the top 20 features by **Gain** — the average improvement in loss brought by a feature.

```{r feature-importance, fig.width=9, fig.height=7}
importance_matrix <- xgb.importance(model = xgb_model)

xgb.plot.importance(
  importance_matrix[1:20, ],
  main = "Top 20 Features by Gain — XGBoost",
  col  = "steelblue"
)
```

**Interpretation**: The features with the highest gain are the strongest drivers of wage prediction. Country of employment typically dominates, followed by age, experience, job role, and education, consistent with real-world wage determinants.

### 5.2 DALEX Explainer Setup

```{r dalex-explainer, message=FALSE, warning=FALSE}
xgb_predict_fun <- function(model, newdata) {
  predict(model, xgb.DMatrix(as.matrix(newdata)))
}

explainer <- DALEX::explain(
  model            = xgb_model,
  data             = as.data.frame(X_test),
  y                = y_test,
  predict_function = xgb_predict_fun,
  label            = "XGBoost",
  verbose          = FALSE
)
```

To interpret our XGBoost model we use DALEX (Descriptive mAchine Learning EXplanations), a model-agnostic explainability framework. DALEX works by wrapping around any trained model. Regardless of its internal complexity and providing a unified set of tools to understand its behaviour. This is particularly valuable for XGBoost, which is inherently a black box model. Without explainability tools we would know what the model predicts but not why. DALEX solves this by providing three complementary techniques:

-   Partial Dependence Profiles: Show how the predicted wage changes as a single feature varies, holding all other features constant. This reveals the direction and shape of each feature's effect on wages globally across all observations.
-   Break-Down plots: Decompose a single prediction into individual feature contributions, showing exactly how each variable pushed the predicted wage up or down from the average baseline.
-   SHAP values: A theoretically rigorous extension of break-down plots that accounts for feature interactions, providing the most accurate attribution of each feature's contribution to a specific prediction.

Together these three techniques give us both a global understanding of the model (which features matter most overall) and a local understanding (why did the model predict this specific wage for this specific person).

### 5.3 Partial Dependence Profiles

PDPs show how the predicted wage changes as a single feature varies, with all other features held constant. This reveals the direction and shape of each feature's effect.

```{r pdp, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
top_features <- importance_matrix$Feature[1:5]

pdp_profiles <- model_profile(
  explainer = explainer,
  variables = top_features,
  type      = "partial"
)

plot(pdp_profiles) +
  labs(title = "Partial Dependence Profiles — Top 5 Features") +
  theme_minimal()
```

### 5.4 SHAP Break-Down for One Prediction

The break-down plot decomposes a single prediction into feature contributions — showing exactly how each variable pushed the predicted wage up or down from the baseline.

```{r shap-breakdown, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
obs_to_explain <- as.data.frame(X_test)[1, , drop = FALSE]

bd <- predict_parts(
  explainer       = explainer,
  new_observation = obs_to_explain,
  type            = "break_down"
)

plot(bd) +
  labs(title = "Break-Down Explanation — Single Prediction") +
  theme_minimal()
```

### 5.5 SHAP Values

SHAP (SHapley Additive exPlanations) values provide a theoretically grounded attribution of each feature's contribution. Unlike break-down plots, SHAP accounts for feature interactions by averaging over all possible feature orderings.

```{r shap-values, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
shap_vals <- predict_parts(
  explainer       = explainer,
  new_observation = obs_to_explain,
  type            = "shap",
  B               = 15
)

plot(shap_vals) +
  labs(title = "SHAP Values — Single Prediction") +
  theme_minimal()
```

**Interpretation**: Positive SHAP values push the prediction above the average wage; negative values push it below. For example, being located in the United States or Switzerland typically adds significantly to the predicted wage, while being a student or having no ML experience reduces it.

------------------------------------------------------------------------

## 6. Wage Prediction for Team Members

We predict each team member's expected wage by constructing their profile using the one-hot encoded feature space. All columns default to 0, and we set the relevant ones to 1 based on each person's actual characteristics.

**Edit the values below to match your real team profiles.**

```{r team-prediction}
# Zero-filled template with all XGBoost features
template_xgb <- as.data.frame(matrix(0, nrow = 1, ncol = ncol(X_train)))
colnames(template_xgb) <- colnames(X_train)

make_xgb_profile <- function(template, settings) {
  profile <- template
  for (col in names(settings)) {
    if (col %in% colnames(profile)) {
      profile[[col]] <- settings[[col]]
    } else {
      warning(paste("Column not found:", col))
    }
  }
  return(profile)
}

# ---- Team Member 1 ---- (edit these!)
member1 <- make_xgb_profile(template_xgb, list(
  age.25.29                          = 1,
  gender.Male                        = 1,
  country.Switzerland                = 1,
  education.Bachelor.s.degree         = 1,
  undergraduate_major.Computer.science..software.engineering..etc.. = 1,
  job_role.Software.Engineer            = 1,
  years_experience.1.2               = 1,
  Programming_Python                 = 1,
  Programming_Java = 1,
  Programming_Javascript.Typescript = 1,
  ML_framework_PyTorch = 1,
ML_framework_Xgboost = 1,
ML_framework_randomForest = 1,
  Programming_R                      = 1,
  Programming_SQL                    = 1,
  ML_atwork.We.recently.started.using.ML.methods..i.e...models.in.production.for.less.than.2.years. = 1,
Visualization_ggplot2 = 1,
Visualization_Matplotlib =1,
Visualization_Shiny = 1,
Visualization_Plotly = 1
))




# ---- Team Member 2 ---- (edit these!)
member2 <- make_xgb_profile(template_xgb, list(
  age.22.24                          = 1,
  gender.Male                      = 1,
  country.Switzerland                    = 1,
  education.Bachelor.s.degree          = 1,
  undergraduate_major.I.never.declared.a.major = 1,
  job_role.Principal.Investigator = 1,
  years_experience.0.1 = 1,
  industry.Military.Security.Defense = 1,
  Programming_Python = 1,
  Programming_R = 1,
  Programming_SQL = 1,
  Programming_Bash = 1
))



# ---- Team Member 3 ---- (edit these!)
member3 <- make_xgb_profile(template_xgb, list(
  age.22.24                          = 1,
  gender.Male                      = 1,
  country.Switzerland                    = 1,
  education.Bachelor.s.degree          = 1,
  undergraduate_major.Computer.science..software.engineering..etc.. = 1,
  job_role.Data.Analyst = 1,
  years_experience.1.2 = 1,
  industry.Computers.Technology = 1,
  Programming_Python = 1,
  Programming_R = 1,
  Programming_SQL = 1,
  Programming_Bash = 1,
  Programming_Javascript.Typescript = 1,
  ML_framework_PyTorch = 1,
  ML_framework_Xgboost = 1,
  ML_framework_randomForest = 1,
  ML_framework_H20 = 1,
  Visualization_ggplot2 = 1,
  Visualization_Shiny = 1,
Visualization_Plotly = 1
))


# Predict
team_profiles   <- bind_rows(member1, member2, member3)
team_dmatrix    <- xgb.DMatrix(as.matrix(team_profiles))
team_log_preds  <- predict(xgb_model, team_dmatrix)
team_wage_preds <- expm1(team_log_preds)

team_results <- data.frame(
  Member         = c("Deniz", "Thiraviyachelvam", "Sen"),
  Predicted_Wage = paste0("$", format(round(team_wage_preds, 0), big.mark = ","))
)

print(team_results)
```

```{r team-plot, fig.width=7, fig.height=4}
team_plot_df <- data.frame(
  Member = c("Deniz", "Thiraviyachelvam", "Sen"),
  Wage   = team_wage_preds
)

ggplot(team_plot_df, aes(x = Member, y = Wage, fill = Member)) +
  geom_col(alpha = 0.85, width = 0.5) +
  geom_text(aes(label = paste0("$", format(round(Wage, 0), big.mark = ","))),
            vjust = -0.5, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Predicted Wages for Team Members",
       x = NULL, y = "Predicted Wage (USD)") +
  theme_minimal() +
  theme(legend.position = "none")

```
```{r shap-member1, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
shap_m1 <- predict_parts(
  explainer       = explainer,
  new_observation = member1,
  type            = "shap",
  B               = 15
)
plot(shap_m1) +
  labs(title = "SHAP Values Deniz") +
  theme_minimal()
```

```{r shap-member2, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
shap_m2 <- predict_parts(
  explainer       = explainer,
  new_observation = member2,
  type            = "shap",
  B               = 15
)
plot(shap_m2) +
  labs(title = "SHAP Values Thiraviyachelvam") +
  theme_minimal()
```

```{r shap-member3, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
shap_m3 <- predict_parts(
  explainer       = explainer,
  new_observation = member3,
  type            = "shap",
  B               = 15
)
plot(shap_m3) +
  labs(title = "SHAP Values Sen") +
  theme_minimal()
```

**Reflection**: The predicted wages reflect population-level patterns from the Kaggle Data Science Survey. Country has the largest effect — Switzerland and Germany consistently predict higher wages than the global average, reflecting higher cost of living and strong data science markets. Experience and education level further refine predictions within each country. These estimates should be treated as benchmarks rather than precise forecasts, as individual factors like company size, negotiation, and specific skills are not captured in the survey.

------------------------------------------------------------------------

## 7. Conclusions

### Model Choice

We selected **XGBoost** as our sole model because it is the best-performing algorithm for tabular data with many sparse binary features. Its gradient boosting mechanism — building trees sequentially to correct previous errors — combined with strong regularisation parameters (`lambda`, `alpha`, `gamma`, `min_child_weight`) makes it particularly well-suited to noisy survey data.

### Model Performance

-   R² = **`r round(xgb_r2, 3)`** on the held-out test set
-   RMSE ≈ **`r round(xgb_rmse_orig, 0)` USD** in original scale
-   MAE ≈ **`r round(xgb_mae_orig, 0)` USD** in original scale

An R² of \~0.31–0.40 is expected and acceptable for this dataset. The primary reasons are: (1) age and experience are bucketed ranges, not exact values; (2) wages are self-reported across vastly different countries and industries; (3) no continuous numeric predictors are available. This R² is consistent with published wage prediction models on similar survey datasets.

### Feature Decisions

-   **Included**: demographics (age, gender, country), education, experience, job role, industry, programming skills, ML tool usage, cloud platforms, activities, coding time percentage, ML experience years
-   **Excluded**: Columns with \>40% missing values
-   **Transformed**: All categoricals one-hot encoded (260+ features); wage log-transformed to address right skew

### Key Drivers of Wage

Based on XGBoost feature importance (Gain) and SHAP values:

-   **Country** — the single strongest predictor, reflecting labour market and cost-of-living differences
-   **Age and years of experience** — strong positive relationship with wage
-   **Job role** — Chief Officers, Software Engineers, and Data Scientists predict higher wages
-   **Education** — Doctoral and Master's degrees associated with higher wages
-   **ML maturity at work** — respondents at companies with established ML in production earn more

### Limitations

-   The model captures survey-level patterns, not individual circumstances
-   Wage brackets are self-reported and may be imprecise
-   The dataset reflects the Kaggle community, which skews towards certain countries and roles
-   Team predictions are benchmarks — actual salaries depend on company, negotiation, and local market conditions
