---
title: "Assignment"
format: html
editor: visual
---

# Predictive modelling for wages

## Objectives

-   Build a model that will predict the variable “wage”. Put differently, the dependent variable of the model should be y = “wage”.

-   You are free to decide on the features you include as explanatory variables as well as the specific models you use. You need to provide details as to decisions you have taken («why did you choose algorithm XYZ»? «How did you choose the features?» «How did you transform them»? Etc.).

-   Once you have built your predictive model, you need to explain it! You need to run explainability techniques so to identify the main features that drive the wage predictions in your model.

-   How will your model predict for you? Use your model to predict the wage that each one from the team will earn in the future.

## Setups and libraries

```{r, Setup}

library(tidyverse)
library(caret)
library(randomForest)
library(xgboost)
library(DALEX)
library(iml)
library(ggplot2)
library(corrplot)
library(fastDummies)
library(Metrics)
library(gridExtra)
library(vip)

load("data_wage.Rdata")


```

## Exploring data

```{r, exploration}
summary(data)

# Check for missing data
missing_summary <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(missing_pct = round(missing_count / nrow(data) * 100, 1)) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_pct))
print(missing_summary)


p1 <- ggplot(data, aes(x = wage)) +
  geom_histogram(fill = "steelblue", bins = 40, alpha = 0.8) +
  labs(title = "Distribution of Wage", x = "Wage", y = "Count") +
  theme_minimal()

p2 <- ggplot(data, aes(x = log1p(wage))) +
  geom_histogram(fill = "darkorange", bins = 40, alpha = 0.8) +
  labs(title = "Distribution of log(Wage + 1)", x = "log(Wage + 1)", y = "Count") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Transformation and clean up

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
data_wage
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
