---
title: "Predictive Modelling for Wages"
format: html
editor: visual
---

In today's data-driven economy, understanding what drives wages has become increasingly important for both individuals and organisations. Wages depend on a wide range of factors such as country, education, job role, experience, and the technical skills a person brings to the table. Capturing these relationships in a single predictive model is challenging, but it provides valuable insights into how the labour market values different profiles.

This project explores wage prediction based on a large global survey of data science professionals from Kaggle, containing 10'809 responses across 78 variables. Our goal is to build a model that predicts an individual's wage based on their demographic, educational, and professional characteristics, and to understand which factors drive these predictions the most.

To achieve this, we apply XGBoost, a gradient boosting algorithm widely recognised as one of the best methods for tabular data. We complement the model with modern explainability techniques such as SHAP values and Partial Dependence Profiles, allowing us to interpret not only what the model predicts, but also why.

The project follows a structured workflow. We begin with an exploratory analysis of the dataset to identify patterns and potential predictors. We then clean and transform the data, train and tune the model using cross-validation, and evaluate its performance on a held-out test set. Finally, we apply the model to our own profiles to predict our expected future wages and reflect critically on the results.

The objectives of this work are threefold: first, to deliver a reliable and well-tuned predictive model; second, to make the model's decisions transparent and interpretable; and third, to honestly discuss the strengths and limitations of survey-based wage prediction in a global context.

## 1. Setup and Libraries

We use `tidyverse` for data manipulation, `caret` for preprocessing, `xgboost` as our primary model, and `DALEX` for explainability.

```{r setup, message=FALSE, warning=FALSE}
packages <- c("tidyverse", "caret", "xgboost",
              "DALEX", "ggplot2", "corrplot",
              "Metrics", "gridExtra")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

library(tidyverse)
library(caret)
library(xgboost)
library(DALEX)
library(ggplot2)
library(corrplot)
library(Metrics)
library(gridExtra)

load("data_wage.Rdata")
```

------------------------------------------------------------------------

## 2. Exploratory Data Analysis

```{r eda, message=FALSE, warning=FALSE}
dim(data)
```

Chech for missing Data.

```{r eda-missing}
missing_summary <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(missing_pct = round(missing_count / nrow(data) * 100, 1)) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_pct))
print(missing_summary)
```

We create a histogram of raw wage values. The second plot are for wages where they are really high and with the help of Log transformation we can now have a better understanding.

```{r eda-wage-dist, fig.width=10, fig.height=4}
p1 <- ggplot(data, aes(x = wage)) +
  geom_histogram(fill = "steelblue", bins = 40, alpha = 0.8) +
  labs(title = "Distribution of Wage", x = "Wage", y = "Count") +
  theme_minimal()

p2 <- data %>%
  filter(!is.na(wage), wage > 0) %>%
  ggplot(aes(x = log1p(wage))) +
  geom_histogram(fill = "darkorange", bins = 40, alpha = 0.8) +
  scale_x_continuous(
    breaks = log1p(c(1000, 10000, 50000, 100000, 200000, 500000)),
    labels = c("$1k", "$10k", "$50k", "$100k", "$200k", "$500k")
  ) +
  labs(title = "Distribution of log(Wage + 1)", 
       x = "Wage (USD)", y = "Count") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The distribution is strongly right-skewed with notable outliers. We apply a log transformation to reduce skew and improve model performance.

```{r eda-categorical, fig.width=10, fig.height=5}
if ("gender" %in% colnames(data)) {
  p3 <- data %>%
  filter(!is.na(gender), !is.na(wage), wage > 0) %>%
  ggplot(aes(x = gender, y = wage, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Wage by Gender (log scale)", x = "Gender", y = "Wage (log scale)") +
  theme_minimal() + theme(legend.position = "none")
  print(p3)
}

if ("education" %in% colnames(data)) {
 p4 <- data %>%
  filter(!is.na(education), !is.na(wage), wage > 0) %>%
  ggplot(aes(x = reorder(education, wage, median), y = wage, fill = education)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  coord_flip() +
  labs(title = "Wage by Education Level (log scale)", x = "Education", y = "Wage (log scale)") +
  theme_minimal() + theme(legend.position = "none")
  print(p4)
}

if ("job_role" %in% colnames(data)) {
  p5 <- data %>%
    filter(!is.na(job_role), !is.na(wage)) %>%
    group_by(job_role) %>%
    summarise(median_wage = median(wage, na.rm = TRUE)) %>%
    ggplot(aes(x = reorder(job_role, median_wage), y = median_wage, fill = median_wage)) +
    geom_col(alpha = 0.85) +
    coord_flip() +
    scale_fill_viridis_c() +
    labs(title = "Median Wage by Job Role", x = "Job Role", y = "Median Wage") +
    theme_minimal() + theme(legend.position = "none")
  print(p5)
}
```

The wage distribution shows slight variation across gender groups, however the differences are too small to draw any meaningful conclusions. Regarding education level, a clear positive relationship is visible, higher academic degrees are consistently associated with higher wages. Finally, looking at wage by job role, Chief Officers, Principal Investigators, and Managers stand out as the highest earning roles, earning significantly more than other positions.

```{r eda-numeric, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
# 1. Top programming languages by usage rate
prog_cols <- data %>%
  select(starts_with("Programming_")) %>%
  select(-Programming_language_used_most_often) %>%
  summarise(across(everything(), ~ mean(. == 1, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "language", values_to = "usage_rate") %>%
  mutate(language = str_remove(language, "Programming_")) %>%
  arrange(desc(usage_rate))

p1 <- ggplot(prog_cols, aes(x = reorder(language, usage_rate), y = usage_rate)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Programming Language Usage", x = NULL, y = "% of Respondents") +
  theme_minimal()

# 2. ML framework usage rate
ml_cols <- data %>%
  select(starts_with("ML_framework_")) %>%
  summarise(across(everything(), ~ mean(. == 1, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "framework", values_to = "usage_rate") %>%
  mutate(framework = str_remove(framework, "ML_framework_")) %>%
  arrange(desc(usage_rate))

p2 <- ggplot(ml_cols, aes(x = reorder(framework, usage_rate), y = usage_rate)) +
  geom_col(fill = "darkorange", alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "ML Framework Usage", x = NULL, y = "% of Respondents") +
  theme_minimal()

# 3. Wage by programming language (Python vs R vs SQL users)
lang_wage <- data %>%
  filter(!is.na(wage)) %>%
  mutate(
    uses_python = ifelse(Programming_Python == 1, "Python", "No Python"),
    uses_r      = ifelse(Programming_R == 1,      "R",      "No R"),
    uses_sql    = ifelse(Programming_SQL == 1,     "SQL",    "No SQL")
  ) %>%
  pivot_longer(cols = c(uses_python, uses_r, uses_sql),
               names_to = "lang_type", values_to = "uses_lang") %>%
  filter(!is.na(uses_lang))

p3 <- ggplot(lang_wage, aes(x = uses_lang, y = wage, fill = uses_lang)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ lang_type, scales = "free_x") +
  labs(title = "Wage by Key Programming Language Usage (log scale)",
       x = NULL, y = "Wage (log scale)") +
  theme_minimal() +
  theme(legend.position = "none")

# 4. Wage by cloud platform usage
cloud_wage <- data %>%
  filter(!is.na(wage)) %>%
  mutate(
    uses_aws   = ifelse(cloud_Amazon.Web.Services..AWS.       == 1, "AWS",   "No AWS"),
    uses_gcp   = ifelse(cloud_Google.Cloud.Platform..GCP.     == 1, "GCP",   "No GCP"),
    uses_azure = ifelse(cloud_Microsoft.Azure                  == 1, "Azure", "No Azure")
  ) %>%
  pivot_longer(cols = c(uses_aws, uses_gcp, uses_azure),
               names_to = "cloud_type", values_to = "uses_cloud") %>%
  filter(!is.na(uses_cloud))

p4 <- ggplot(cloud_wage, aes(x = uses_cloud, y = wage, fill = uses_cloud)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ cloud_type, scales = "free_x") +
  labs(title = "Wage by Cloud Platform Usage (log scale)",
       x = NULL, y = "Wage (log scale)") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The relationship between programming language usage and wages is not immediately clear, making it difficult to draw strong conclusions from the data. However, cloud platform usage tells a different story, there is a notable difference in wages between users and non-users across all three platforms. In particular, AWS users appear to earn significantly more than non-users, suggesting that cloud platform experience, especially AWS, is associated with higher compensation. This may reflect the fact that cloud skills are in high demand in the industry, commanding a wage premium.

------------------------------------------------------------------------

## 3. Data Transformation and Cleaning

### 3.1 Feature Selection Rationale

We select features based on three criteria:

1.  **Domain relevance**: Variables that intuitively drive wages (experience, education, skills, role, country)
2.  **Data quality**: Dropping columns with \>40% missing values
3.  **Avoiding data leakage**: Excluding any variable that directly encodes wage

Wages are not internationally uniform, compensation in prosperous economies is significantly higher than in developing countries, making country of employment one of the most important variables in our transformation and modelling process. As part of our data quality checks, we also screened for missing values. We applied a 40% threshold, meaning any variable with more than 40% missing values was removed. This ensures we only retain features that carry meaningful information. Finally, to prevent data leakage, we carefully excluded any variables that directly encode wage. Including such variables would artificially inflate model performance, as the model would essentially be using the answer to predict the answer, producing misleadingly optimistic results that would not generalise to new data.

```{r cleaning, message=FALSE, warning=FALSE}
# Step 1: Remove rows where wage is missing
data_clean <- data %>% filter(!is.na(wage))
cat("Rows after removing missing wage:", nrow(data_clean), "\n")

# Step 2: Drop columns with >40% missing values
missing_threshold <- 0.40
keep_cols <- colnames(data_clean)[colMeans(is.na(data_clean)) <= missing_threshold]
data_clean <- data_clean %>% select(all_of(keep_cols))
cat("Columns remaining after missing filter:", ncol(data_clean), "\n")

# Step 3: Log-transform the target (wage is right-skewed)
data_clean <- data_clean %>%
  mutate(log_wage = log1p(wage)) %>%
  select(-wage)

# Step 4: Handle remaining NAs
data_clean <- data_clean %>%
  mutate(across(where(is.character), ~ replace_na(., "Unknown"))) %>%
  mutate(across(where(is.factor),   ~ fct_explicit_na(., na_level = "Unknown"))) %>%
  mutate(across(where(is.numeric),  ~ replace_na(., median(., na.rm = TRUE))))

cat("Final clean dataset dimensions:", dim(data_clean), "\n")
```

```{r encoding}
# Step 5: Encode character columns as factors
cat_vars <- data_clean %>% select(where(is.character)) %>% colnames()
data_model <- data_clean %>%
  mutate(across(all_of(cat_vars), as.factor))
```

### 3.2 Train / Test Split

We apply an 80/20 train/test split, allocating 80% of the data for training and 20% for evaluation. Stratification on log_wage ensures that both sets have a representative wage distribution, preventing the model from being evaluated on an unrepresentative sample. This split ratio was selected after experimenting with alternative configurations, including 70/30 and 90/10 splits, and found to offer the best balance between having sufficient training data for the model to learn from and enough test data to produce reliable performance estimates.

```{r split}
set.seed(42)

train_index <- createDataPartition(data_model$log_wage, p = 0.8, list = FALSE)
train_data  <- data_model[train_index, ]
test_data   <- data_model[-train_index, ]

cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:    ", nrow(test_data),  "\n")
```

### 3.3 One-Hot Encoding for XGBoost

XGBoost requires fully numeric input. We one-hot encode all factor columns using `dummyVars`, resulting in 260+ binary features.

```{r ohe}
dummies   <- dummyVars(log_wage ~ ., data = train_data)
train_xgb <- predict(dummies, newdata = train_data) %>% as.data.frame()
train_xgb$log_wage <- train_data$log_wage

test_xgb  <- predict(dummies, newdata = test_data) %>% as.data.frame()
test_xgb$log_wage  <- test_data$log_wage

# Clean column names
names(train_xgb) <- make.names(names(train_xgb), unique = TRUE)
names(test_xgb)  <- make.names(names(test_xgb),  unique = TRUE)

# XGBoost matrices
X_train <- as.matrix(train_xgb %>% select(-log_wage))
y_train <- train_xgb$log_wage
X_test  <- as.matrix(test_xgb  %>% select(-log_wage))
y_test  <- test_xgb$log_wage

dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

cat("Total features after encoding:", ncol(X_train), "\n")
```

------------------------------------------------------------------------

## 4. Choosing a model which makes sense

We tested several models to identify the most suitable approach for our dataset, including Linear Regression, Random Forest, and XGBoost.

Linear Regression was the first model we evaluated. However, it proved to be a poor fit for this problem. Linear Regression assumes straight-line relationships between features and the target variable, treating every feature's effect on wage as constant and additive. In reality, wages are far more complex. A Data Scientist in Switzerland with 5 years of experience earns disproportionately more than the sum of those individual factors would suggest. Real-world compensation is influenced by interactions between country, role, experience, and education that a linear model simply cannot capture. Wages do not follow a fixed rulebook, and assuming they do leads to systematic prediction errors.

Random Forest was the second model we evaluated. While it performed significantly better than Linear Regression, it came with several weaknesses that made it unsuitable for our case.

The most significant issue was computation time. Training the Random Forest took approximately 25 minutes, which was far too slow for our workflow and made iterating and testing different configurations impractical.

Additionally, our dataset contains 260+ binary features, the majority of which are zero for any given respondent. This proved to be a fundamental problem for Random Forest. This combination of poor computational efficiency and difficulty handling sparse binary data made Random Forest a less suitable choice compared to XGBoost, which handles both of these issues natively.

XGBoost builds trees sequentially, meaning each new tree learns from the mistakes of the previous one. This boosting mechanism allows the model to progressively improve its predictions with every iteration, resulting in significantly higher accuracy compared to our previous models.

XGBoost also handles our 260+ sparse binary features considerably better than Random Forest. Its built-in mechanism for dealing with zero values ensures that uninformative features do not weaken the model, which was a critical issue we encountered with Random Forest.

Furthermore, XGBoost was significantly faster in terms of computation time. As it allowed us to test multiple configurations and tune hyperparameters efficiently without the 25-minute waiting times we experienced with Random Forest.

Finally, we discovered that XGBoost offers native SHAP compatibility, which greatly simplified our explainability analysis. This allowed us to not only predict wages accurately but also gain a clear understanding of which features drive each individual prediction a key requirement of this assignment.

### 4.1 Hyperparameter Tuning

While XGBoost is a powerful algorithm, its default settings are not necessarily optimal for every dataset. This is why we tuned its hyperparameters.

Through a process of trial and error we tested multiple parameter combinations using 5-fold cross-validation, evaluating each configuration on unseen data to ensure we selected settings that generalise well rather than simply memorising the training data. The parameters we tuned and their final values are:

-   **`eta` (0.05)**: The learning rate. Controls how much each tree contributes to the final prediction. A lower value of 0.05 means the model learns slowly and precisely, reducing the risk of overshooting the optimal solution at the cost of requiring more boosting rounds.

-   **`max_depth` (5)**: The maximum depth of each tree. Controls model complexity. A depth of 5 balances the ability to capture complex wage patterns without overfitting to noise in the survey data.

-   **`subsample` (0.8)**: Each tree is trained on a random 80% of the training rows. Introducing randomness that prevents trees from becoming too similar to each other and improves generalisation.

-   **`colsample_bytree` (0.7)**: Each tree sees a random 70% of features. Particularly useful for our dataset where many of the 260+ binary features are correlated with each other.

-   **`min_child_weight` (5)**: The minimum number of observations required in a leaf node. Prevents the model from learning patterns that only apply to very small groups of respondents, which are likely noise rather than real wage signals.

-   **`gamma` (0.1)**: The minimum loss reduction required to make a split. Acts as a pruning mechanism, ensuring the model only makes splits that meaningfully improve predictions.

```{r xgboost-tune, message=FALSE, warning=FALSE, cache=TRUE}
# Best parameters found through trial and error
best_params <- list(
  objective        = "reg:squarederror",
  eval_metric      = "rmse",
  eta              = 0.05,
  max_depth        = 5,
  min_child_weight = 3,
  gamma            = 0.0,
  subsample        = 0.8,
  colsample_bytree = 0.7,
  lambda           = 1.0,
  alpha            = 0.0
)

# Use cross-validation to find optimal number of rounds
set.seed(42)
xgb_cv <- xgb.cv(
  data    = dtrain,
  nrounds = 600,
  nfold   = 5,
  params  = best_params,
  verbose = FALSE
)

best_nrounds <- which.min(xgb_cv$evaluation_log$test_rmse_mean)
cat(sprintf("Best number of rounds: %d\n", best_nrounds))

# Train final model
set.seed(42)
xgb_model <- xgb.train(
  data    = dtrain,
  nrounds = best_nrounds,
  params  = best_params,
  verbose = 0
)

cat("Model training complete.\n")
```

### 4.2 Model Evaluation

```{r, xgboost-eval, message=FALSE, warning=FALSE, cache=TRUE}
# Predictions
xgb_preds_log  <- predict(xgb_model, dtest)
xgb_preds_orig <- expm1(xgb_preds_log)
actual_orig    <- expm1(y_test)

# Metrics
xgb_r2        <- cor(y_test, xgb_preds_log)^2
xgb_rmse_orig <- rmse(actual_orig, xgb_preds_orig)
xgb_mae_orig  <- mae(actual_orig,  xgb_preds_orig)

# Results table
data.frame(
  Metric = c("R²", "RMSE (USD)", "MAE (USD)"),
  Value  = c(round(xgb_r2, 3),
             round(xgb_rmse_orig, 0),
             round(xgb_mae_orig,  0))
)
```

Initially, the XGBoost model produced a lower than expected R² of approximately 0.33. During our investigation we discovered that excluding wages equal to zero nearly doubled the model's accuracy, bringing R² to approximately 0.65. After careful consideration however, we decided to retain the zero-wage observations in our final model. While their removal improves performance metrics, excluding them introduces a selection bias, students, unemployed respondents, and those who chose not to disclose their income are a real and meaningful part of the population. Removing them would make the model less representative of the full workforce and could lead to misleading predictions for profiles that resemble these groups.

```{r eval-plots, fig.width=10, fig.height=4}
residuals_df <- data.frame(
  actual    = actual_orig,
  predicted = xgb_preds_orig,
  residual  = actual_orig - xgb_preds_orig
)

# Actual vs Predicted
p1 <- ggplot(residuals_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue", size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted", x = "Actual (USD)", y = "Predicted (USD)") +
  theme_minimal()

# Residual distribution
p2 <- ggplot(residuals_df, aes(x = residual)) +
  geom_histogram(fill = "steelblue", bins = 40, alpha = 0.8) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residual Distribution", x = "Residual (USD)", y = "Count") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

To evaluate model performance visually we produce three diagnostic plots. The first plot shows Actual vs Predicted Wage, in a perfect model all points would fall exactly on the red dashed line. Points scattered around it indicate prediction errors, with points above the line meaning the model underestimated and points below meaning it overestimated. The second plot shows Residuals vs Predicted, residuals are the difference between actual and predicted wages. Ideally these should be randomly scattered around zero with no clear pattern. A pattern would suggest the model is systematically wrong for certain wage ranges. The third plot shows the Distribution of Residuals, ideally this should be a bell curve centered at zero, meaning errors are symmetric and the model is not consistently over or under predicting. In our case the residuals are approximately centered at zero, which is a good sign, though the spread reflects the inherent difficulty of predicting wages from survey data.

The model achieves an R² of approximately 0.33, which while modest is expected and acceptable for this dataset. Age and experience are bucketed ranges rather than exact values, wages are self-reported across vastly different countries and roles, and almost all features are categorical, leaving the model with limited ability to capture fine-grained patterns. Survey respondents may also approximate or misreport their salaries, adding further noise. An R² of 0.33 on a noisy global wage survey is consistent with published research on similar datasets.

------------------------------------------------------------------------

## 5. Model Explainability

### 5.1 Feature Importance

XGBoost's built-in importance measures show which features contributed most to splits across all trees. We show the top 20 features by **Gain**

```{r feature-importance, fig.width=9, fig.height=7}
importance_matrix <- xgb.importance(model = xgb_model)

xgb.plot.importance(
  importance_matrix[1:20, ],
  main = "Top 20 Features by Gain"
)
```

**Interpretation**: The features with the highest gain are the strongest drivers of wage prediction. Country of employment typically dominates, followed by age, experience, job role, and education, consistent with real-world wage determinants.

### 5.2 DALEX Explainer Setup

```{r dalex-explainer, message=FALSE, warning=FALSE}
xgb_predict_fun <- function(model, newdata) {
  predict(model, xgb.DMatrix(as.matrix(newdata)))
}

explainer <- DALEX::explain(
  model            = xgb_model,
  data             = as.data.frame(X_test),
  y                = y_test,
  predict_function = xgb_predict_fun,
  label            = "XGBoost",
  verbose          = FALSE
)
```

To interpret our XGBoost model we use DALEX (Descriptive mAchine Learning EXplanations), a model-agnostic explainability framework. DALEX works by wrapping around any trained model. Regardless of its internal complexity and providing a unified set of tools to understand its behaviour. This is particularly valuable for XGBoost, which is inherently a black box model. Without explainability tools we would know what the model predicts but not why. DALEX solves this by providing three complementary techniques:

-   Partial Dependence Profiles: Show how the predicted wage changes as a single feature varies, holding all other features constant. This reveals the direction and shape of each feature's effect on wages globally across all observations.
-   Break-Down plots: Decompose a single prediction into individual feature contributions, showing exactly how each variable pushed the predicted wage up or down from the average baseline.
-   SHAP values: A theoretically rigorous extension of break-down plots that accounts for feature interactions, providing the most accurate attribution of each feature's contribution to a specific prediction.

Together these three techniques give us both a global understanding of the model (which features matter most overall) and a local understanding (why did the model predict this specific wage for this specific person).

### 5.3 Partial Dependence Profiles

PDPs show how the predicted wage changes as a single feature varies, with all other features held constant. This reveals the direction and shape of each feature's effect.

```{r pdp, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
top_features <- importance_matrix$Feature[1:5]

pdp_profiles <- model_profile(
  explainer = explainer,
  variables = top_features,
  type      = "partial"
)

plot(pdp_profiles) +
  labs(title = "Partial Dependence Profiles — Top 5 Features") +
  theme_minimal()
```

### 5.4 SHAP Break-Down for One Prediction

The break-down plot decomposes a single prediction into feature contributions — showing exactly how each variable pushed the predicted wage up or down from the baseline.

```{r shap-breakdown, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
obs_to_explain <- as.data.frame(X_test)[1, , drop = FALSE]

bd <- predict_parts(
  explainer       = explainer,
  new_observation = obs_to_explain,
  type            = "break_down"
)

plot(bd) +
  labs(title = "Break-Down Explanation Single Prediction") +
  theme_minimal()
```

### 5.5 SHAP Values

SHAP (SHapley Additive exPlanations) values provide a theoretically grounded attribution of each feature's contribution. Unlike break-down plots, SHAP accounts for feature interactions by averaging over all possible feature orderings.

```{r shap-values, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
shap_vals <- predict_parts(
  explainer       = explainer,
  new_observation = obs_to_explain,
  type            = "shap",
  B               = 15
)

plot(shap_vals) +
  labs(title = "SHAP Values Single Prediction") +
  theme_minimal()
```

**Interpretation**: Positive SHAP values push the prediction above the average wage; negative values push it below. For example, being located in the United States or Switzerland typically adds significantly to the predicted wage, while being a student or having no ML experience reduces it.

------------------------------------------------------------------------

## 6. Wage Prediction for Team Members

We predict each team member's expected wage by constructing their profile using the one-hot encoded feature space. All columns default to 0, and we set the relevant ones to 1 based on each person's actual characteristics.

**Edit the values below to match your real team profiles.**

```{r team-prediction}
# Zero-filled template
template_xgb <- as.data.frame(matrix(0, nrow = 3, ncol = ncol(X_train)))
colnames(template_xgb) <- colnames(X_train)

# ---- Member 1: Deniz ----
template_xgb["1", "age.25.29"] <- 1
template_xgb["1", "gender.Male"] <- 1
template_xgb["1", "country.Switzerland"] <- 1
template_xgb["1", "education.Bachelor.s.degree"] <- 1
template_xgb["1", "job_role.Software.Engineer"] <- 1
template_xgb["1", "years_experience.1.2"] <- 1
template_xgb["1", "Programming_Python"] <- 1
template_xgb["1", "Programming_R"] <- 1
template_xgb["1", "Programming_SQL"] <- 1
template_xgb["1", "ML_framework_PyTorch"] <- 1

# ---- Member 2: Thiraviyachelvam ----
template_xgb["2", "age.22.24"] <- 1
template_xgb["2", "gender.Male"] <- 1
template_xgb["2", "country.Switzerland"] <- 1
template_xgb["2", "education.Bachelor.s.degree"] <- 1
template_xgb["2", "job_role.Principal.Investigator"] <- 1
template_xgb["2", "years_experience.0.1"] <- 1
template_xgb["2", "Programming_Python"] <- 1
template_xgb["2", "Programming_R"] <- 1
template_xgb["2", "Programming_SQL"] <- 1

# ---- Member 3: Sen ----
template_xgb["3", "age.22.24"] <- 1
template_xgb["3", "gender.Male"] <- 1
template_xgb["3", "country.Switzerland"] <- 1
template_xgb["3", "education.Bachelor.s.degree"] <- 1
template_xgb["3", "job_role.Data.Analyst"] <- 1
template_xgb["3", "years_experience.1.2"] <- 1
template_xgb["3", "Programming_Python"] <- 1
template_xgb["3", "Programming_R"] <- 1
template_xgb["3", "Programming_SQL"] <- 1
template_xgb["3", "ML_framework_PyTorch"] <- 1

# Predict
team_dmatrix    <- xgb.DMatrix(as.matrix(template_xgb))
team_log_preds  <- predict(xgb_model, team_dmatrix)
team_wage_preds <- expm1(team_log_preds)

data.frame(
  Member         = c("Deniz", "Thiraviyachelvam", "Sen"),
  Predicted_Wage = paste0("$", format(round(team_wage_preds, 0), big.mark = ","))
)
```

```{r team-plot, fig.width=7, fig.height=4}
data.frame(
  Member = c("Deniz", "Thiraviyachelvam", "Sen"),
  Wage   = team_wage_preds
) %>%
  ggplot(aes(x = Member, y = Wage, fill = Member)) +
  geom_col(alpha = 0.85, width = 0.5) +
  geom_text(aes(label = paste0("$", format(round(Wage, 0), big.mark = ","))),
            vjust = -0.5, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Predicted Wages for Team Members",
       x = NULL, y = "Predicted Wage (USD)") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r shap-member1, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
shap_m1 <- predict_parts(
  explainer       = explainer,
  new_observation = template_xgb[1, ],
  type            = "shap",
  B               = 15
)

bd1 <- predict_parts(
  explainer       = explainer,
  new_observation = template_xgb[1, ],
  type            = "break_down"
)

plot(bd1) +
  labs(title = "Break-Down Explanation Deniz") +
  theme_minimal()

plot(shap_m1) +
  labs(title = "SHAP Values Deniz") +
  theme_minimal()

shap_m2 <- predict_parts(
  explainer       = explainer,
  new_observation = template_xgb[2, ],
  type            = "shap",
  B               = 15
)

bd2 <- predict_parts(
  explainer       = explainer,
  new_observation = template_xgb[2, ],
  type            = "break_down"
)

plot(bd2) +
  labs(title = "Break-Down Explanation Thiraviyachelvam") +
  theme_minimal()

plot(shap_m2) +
  labs(title = "SHAP Values Thiraviyachelvam") +
  theme_minimal()

shap_m3 <- predict_parts(
  explainer       = explainer,
  new_observation = template_xgb[3, ],
  type            = "shap",
  B               = 15
)

bd3 <- predict_parts(
  explainer       = explainer,
  new_observation = template_xgb[3, ],
  type            = "break_down"
)

plot(bd2) +
  labs(title = "Break-Down Explanation Sen") +
  theme_minimal()

plot(shap_m3) +
  labs(title = "SHAP Values Sen") +
  theme_minimal()
```

By examining the SHAP plots we can gain a clear understanding of how the model makes its wage predictions. A few features stand out as particularly strong influences. Most notably country and age. Most other features, including programming languages and even education level, appear to have relatively small individual effects and are often grouped together in the residual 'other factors' category.
This pattern can be explained in two ways. First, outliers in the data may still be heavily influencing predictions despite our cleaning steps. Second, features like programming languages and education, while intuitively important, may not carry enough signal in this dataset to dominate the prediction individually.
This does not mean the results are meaningless. It makes intuitive sense that a young person with little experience earns significantly less than someone with many years of experience living in a high-income country. However, the SHAP analysis also reveals the limitations of the dataset itself. Important wage drivers such as company size, local market conditions, and individual salary negotiations are not captured in the survey data. SHAP can only reflect what the data provides and in this case it essentially reflects average wages per country.
It is worth noting that if the model were trained on data from a single country, predictions would likely be considerably more accurate, as country-level wage variance would be eliminated. This is both a self-critique of the dataset and a reflection of the modelling choices we made. Nevertheless, within the context of this project, the model provides a solid and interpretable baseline for predicting wages on a global scale.

------------------------------------------------------------------------

# 7 Conlusion

In this project we built a predictive model for wages using survey data from the Kaggle Data Science community. After testing three models, Linear Regression, Random Forest and XGBoost, we selected XGBoost as our final model based on its superior performance, computational efficiency and native SHAP compatibility.
The model achieved an R² of approximately 0.33. While modest, this is expected given that wages are self-reported, globally diverse and influenced by factors not captured in the survey such as company size, local market conditions and individual negotiations. An interesting finding during our process was that excluding zero wages nearly doubled model accuracy, highlighting an important trade-off between predictive performance and representativeness of the full population.
The explainability analysis showed that country and age are by far the strongest wage predictors, which aligns with real-world intuition. Programming languages and education showed relatively small individual effects. Not a failure of the model, but a reflection of the data available.
For the team predictions, the model provided reasonable estimates for each member. Switzerland consistently predicted higher wages, reflecting its strong labour market and high cost of living.
Looking ahead, the most impactful improvement would be richer data. Variables such as company size, city, years in current role and actual negotiated salaries would give the model considerably more signal. Building separate country-specific models would also eliminate the enormous wage variance introduced by combining vastly different labour markets into a single prediction.
Overall this project gave us a solid understanding of the full machine learning pipeline from data cleaning and feature engineering through model selection, hyperparameter tuning and explainability. The model serves as a good baseline for global wage prediction, and with richer and more representative data, accuracy could be improved significantly.
